\chapter{Full Master Narrative}


I began my military career in 2003 when I entered the Cadet School. After 21 months of training, I was commissioned as an Infantry Officer in the 3rd Infantry Battalion at Stephens Barracks in Kilkenny. My career path from the outset was shaped not only by military service but also by academic choices. One of the most decisive was my decision to transfer from a Bachelor of Arts to a Bachelor of Science in Physics. That choice led me to complete a Master’s by Research in Physics. At the time, I had no intention of pursuing a career in the Ordnance Corps --- in fact, I never wanted to become an Ordnance Officer --- but this academic trajectory opened that door.

At around 22 years of age I commenced a period of rapid maturing. I complete the Gold Gaisce (President's Award), began a self-taught audio course in German, began travelling and learning. This happened after a period of significant anti-social behaviour, which was exceptionally formative. Hence, the period 21 - 23 years of age basically changed my life. For the better.

Early postings included service in the External Education and Training Branch in Defence Forces Headquarters. I later completed the Ordnance Young Officers Course, qualifying as an Ordnance Technical Officer in 2016. From there, I took on a series of technical and instructional appointments in the Ordnance School and the Ordnance Company. Operational service took me overseas to Lebanon in 2012 and later to Syria as a Bomb Disposal Officer. Syria in particular left a deep impression on me, as the chaos and nihilism of conflict showed that even when you put in the effort and do the right thing, outcomes can remain illogical or be undone by enemy action.

I also served as Technical Officer in the Army Ranger Wing before returning to Defence Forces Headquarters in the Operations Branch. These experiences gave me a mix of operational, technical, and staff exposure at different levels of the organisation.

A key transformation point in my career came when I was appointed as a Company Commander overseas. During my promotion interview, I had argued that the hardest part of higher command would be managing officers, not soldiers. The board disagreed, but I was later proved right. As a Company Commander, I had officers under my command who ranged from highly competent to completely incompetent, with many disillusioned, sceptical, or disengaged. Managing them was particularly challenging: some needed close supervision, others needed space, and all required something different from me as their leader. It was an intensely difficult but formative experience, teaching me that command at that level is less about tactics and more about balancing personalities, expectations, and human complexity.

Reflecting on my development, I can see how my leadership style has evolved since I was a young Lieutenant. At that stage, I relied heavily on corporals and sergeants and felt obliged to intervene in every small barracks transgression. I believed it was my duty to correct every issue. Today, I no longer see it that way. I speak far less about minor matters, preferring to focus on what is truly important. Silence and restraint, I now understand, can be a sign of maturity. Still, I recognise the risk: sometimes I remain silent when I probably should address an issue.

My leadership style has become situational. I adapt how I lead depending on the group and the context, not because I am inconsistent but because I believe different parts of my personality are appropriate in different circumstances. I avoid micromanagement, preferring to give broad guidance and allow autonomy. I encourage initiative, support learning through experience, and take satisfaction in seeing people grow as a result of my input. I make a deliberate effort to treat subordinates with respect and politeness. That said, I recognise that what I consider support is not always perceived that way. Some expect blind agreement with their assessments or their treatment of others, and when I do not provide that, I have been accused of hypocrisy.

I see myself as both a team-oriented and transformational leader. I actively seek out talent, try to develop colleagues, and value differences among people. I place importance on putting forward a vision and aligning others with it. I design and redesign jobs to make them meaningful and challenging, believing that people should find purpose in their work. Yet I also see myself as situational: I change how I act to match the circumstances, whether leading a technical unit, a company, or a staff group.

The leaders I admire most are those who embody steadiness under pressure. Figures like Major Dick Winters and Major Hans von Luck inspire me because they were cool, calm, collected, and competent under fire. They were not necessarily charismatic or flamboyant, but people respected them and turned to them in crises. That is the style I aspire to: I want to be someone my colleagues and subordinates can turn to in difficult moments, knowing that my input will be dependable and competent. In my specific professional domain, I want to be recognised as the go-to expert in improvised explosive device disposal (IEDD), the trusted authority within the Defence Forces on that subject.

My Emergenetics profile reinforces these tendencies. I am highly Analytical (43 per cent, 95th percentile) and Conceptual (35 per cent, 77th percentile), with much lower preferences for Structural (6 per cent, 13th percentile) and Social (16 per cent, 34th percentile) thinking. Behaviourally, I am extremely Assertive (95th percentile), very low in Expressiveness (15th percentile), and moderate in Flexibility (51st percentile). This profile makes me rational, visionary, forceful, and direct, but often reserved in larger groups. It explains why I thrive on complex, abstract problems but sometimes neglect detail or relational nuance. It also suggests why people may find me intimidating: I tend to drive hard for logical, big-picture solutions, often in a reserved way, but with assertive force when I believe I am right.

My strengths are clear. I am tenacious, able to see tasks through to completion. I have strong staff skills and am regarded as candid and intelligent. My public communication is effective, and I possess high empathy. I take pride in seeking out talent, valuing differences, and helping others to develop.

Yet my weaknesses are just as evident. I have a temper and can be quite angry. I am disorganised, often taking on too much at once and spreading myself too thin. My assertiveness, while a strength, can also make me overbearing; I often struggle to let things go. I am less expressive than most, sometimes too reserved in groups. My social battery is low: I enjoy being around people but quickly need time alone, which can create stress during deployments. I can be cynical, distrusting both individuals and the system. I want to believe in the best in people, but experience has taught me otherwise. I believe strongly in public service, but when institutions fall short, my cynicism grows. I also worry about my resilience in the face of chaos: in the nihilism of war, where logic and fairness no longer apply, I fear I could be affected deeply.

A transformational experience was winning a redress of wrongs against the organisation. It revealed the institution at its worst --- bureaucratic, cynical, and unjust --- but also taught me that fairness and integrity can be defended, even against the system itself. It reinforced my conviction that leaders must sometimes take on their own organisations if they are to uphold the values they claim to represent.

As I begin the Joint Command and Staff Course, my personal development goals are straightforward but important. I want to achieve a grade that reflects my education and input, but not at the expense of my family. Family is now my highest priority. In the past, I did not live that way, but now I intend to. I want this course to enhance my professional growth without undermining my home life. I also want to maintain my fitness throughout the year, exercising three times a week and keeping my body mass steady.

My feelings about the course are mixed. I expect it to be a major burden --- ``a gigantic pain'' --- particularly as I already hold two Master’s degrees and do not have an academic itch to scratch. I worry about the impact on my mental health, personal time, and physical fitness. I fear that the workload will encroach on my home life and undo the balance I have worked hard to create. Another worry is ideological: I am concerned about exposure to radical leftist perspectives within the university environment. My assertiveness makes it difficult for me to stay silent if I feel something is being imposed unfairly. If I speak up, it could cause conflict; if I do not, it will gnaw at me from the inside.

In terms of values, family is paramount. I am determined to prioritise it over everything else, even though in the past I did not. I also remain committed to the values of military and public service, though my cynicism has sometimes eroded my trust in institutions. I strive to always do my best, which makes it difficult for me to consciously lower the bar for this course --- even though I know doing so might protect family time.

In sum, my journey has been defined by technical expertise, operational experience, leadership challenges, and personal evolution. From the Cadet School to bomb disposal in Syria, from the Army Ranger Wing to Defence Forces Headquarters, from youthful micromanagement to situational leadership, from being wronged by the organisation to standing my ground and winning a redress --- all of these experiences have shaped the leader I am today. Now, my focus is on reconciling professional ambition with personal values, particularly family, while striving to embody the calm, competent, and respected leadership that I admire in others.

I thought the presentation was good, a lot of effort involved in it. It was worth doing, I think. I was working with Finola, we worked well together, no issues there. It's good to get back to public speaking, but did it actually teach me anything? Probably not. The reality is, you know, it's probably wasted work, to be honest, really. I don't think I learned a whole lot doing it. I expect I might change my views when I get my results, but yeah, that's how I feel at the moment. The Cranfield University Week with Dr. Dennis Vincent. Okay, you know, high-quality academic stuff, don't get me wrong, and I don't knock his credentials or even the requirement to have some of that deliberate stuff, but I'm not sure how much of it was actually new. It's a lot of fluff, to be honest. Certainly even the leadership stuff, if any of that stuff is new to you at this point in your career, I don't know what the fuck you're doing here, to be honest. But yeah, the models that he implemented were decent, or he introduced us were decent, like the seven S's, then East, and the force field model. Yeah, things like that were, they were okay. Yeah, they were okay, you know, to introduce us to them, but I thought the best part was the last day when we talked about ethics. That was a real discussion, but like, as far as I'm concerned, the, you know, exercise that we're doing during the week were shite by comparison, but we didn't get to do any real discussion on the ethics piece, because we ran out of time. I thought the strategic communications, outsourced strategic communications piece was a disgrace. They were terrible, got nothing out of it. They didn't critique us at all. It was completely bullshit. And I thought the emergenetics personality stuff, that was useful. It was interesting. I'm not sure it was useful, but it's interesting. I also think I was shocked when people were shocked that it gave them, it seems at least one individual, he didn't know any of these things about himself, and I thought that was bizarre, but there you go, it taught me nothing new about myself. If anything, the day was a great way to work as an icebreaker amongst the class. Yeah, Gavin Egerton's leadership mission command thing was, it was interesting, but I'm not sure how much it is really applicable in our organization. Probably the thing that I learned about mission command is that, you know, a leader has to get involved at some point at the lowest level, and people can interpret that as micromanaging, but if you're doing it correctly, you're still getting involved. Doing the literature or the dissertation proposal was a gigantic pain in the hole. Fucking gigantic pain in the hole. Brian and I can meet together and obviously we end up discussing all sorts of things regarding the course. Discussing all sorts of things regarding the presentations as well, and his one was on diversity, and we ended up having debates about diversity within the Defence Forces and generally. And I noticed that Brian's views are very sympathetic and empathetic, and mine are less so.

\chapter{DSS}
% Approach (qual/quant/mixed), cases + justification, data & analysis,
% limits, ethics. Social-science or humanities angle as per guideline.

Reflective Journal – DSS Module

Tuesday – Day 2

Day 1, I came home quite disheartened, feeling that we were, once again, just learning information for the sake of it, not really, you know, let’s do a master’s for the sake of it.

Today then, we were talking, we had Rory Finnegan primarily, and somebody else then from DCU talking about Realpolitik and the realist and liberal approaches and the neoliberalism, and, sorry, neorealism, excuse me, not neoliberalism, and I thought it was interesting, but it wasn’t going any place for me.

And then on the way home, I was talking with Brian and Liam, and I was saying that I was reading Colin Gray’s 2018 book on policy, sorry, on strategy, and I think I was saying that I was understanding that strategy is about consequences, political consequences. Operations are about the movement of resources and whatnot to achieve what is done at the tactical level. The tactical level is where the killing is done, and that can be anywhere by any amount of forces, but ultimately, operations are about putting those people into the position where the killing can happen at the tactical level.

And then he was talking about ends, ways and means, and I think ends are the political piece, that’s the policy. The ways are, I think, the strategy to achieve those, which are long-term, you might never conceive them, and the means are the tactics, the actual tactical outputs that you’re having.

Then I was thinking that, I still was thinking that, and all of this is kind of nonsense really, that it’s going nowhere, what’s the bloody point of listening to it? And then just reading for something else, I was reading Colin Gray’s book, sorry, I was reading Eliot Cohen’s 2002 paper, Eliot Cohen’s 2002 paper called, I can’t remember exactly what it’s called, but any of these papers on about the friction, the friction between the civilians and the military, and that there has to be friction in your job, and he mentioned the operational level, and I was thinking then, trying to square that with Gray’s book, saying well actually, you shouldn’t just be mouthing off at the operational level, like you still don’t understand where the military guys fit in with the strategy, and then I realised that the military strategic thinkers, they have to be, they have to understand the geopolitics, the economics, the culture, all these sorts of things, so that they can influence the political policy, and thereafter can come up with probably, and that’s at maybe the cabinet level, maybe, or maybe at parliament level, and then at the operational level, which might be them engaging with the Department of Defence, that’s definitely frictional, definitely a bounce back of ideas and whatnot, and you just can’t do it effectively unless you know about them.

Then I was reading Gray’s 2005 paper, How Has War Changed Since the End of the Cold War, and he’s on about, I don’t know what he’s on about, but that’s what I have at the moment.and.

Then, the nature of war is about being politics by other means. It's about effecting policy using military tools. Whereas the character of war is something I suppose I'd be more familiar with. GRAY_2018 AND COHEN_2002	


Perfect — here’s your cleaned transcript of what you’ve said so far, presented as a single block of raw text (tidied for readability, but still mostly verbatim). You can copy and paste this straight into Word.

⸻

Transcript of Week 2 Notes (DSS / Maynooth Block)

So, we just finished week 2 of the DSS — I think that’s Defence and Security Studies — block. They call it the Maynooth block, spelled M-A-Y-N-O-O-T-H. Last week was in Maynooth fully, and this week, due to accommodation issues, we were all remoting from home.

This week I was busy trying to get my DSS essay sourced, getting the things I need. As a result, I spent most of the week using GPT to help identify a structure and prolific authors. Then I went on to the various article linking sites to map out my articles, map out the authors, and find the articles I wanted. I then went into a black hole of looking at articles.

I made a few GPT prompts to analyse my sources and extract data. I started using the matrix that Angela had suggested, so my GPT extracts the matrix data and produces a DIMER output, which I put into a master sheet of partially analysed source data. From that I get GPT to produce a draft skeleton of an essay, based on the structure I give it. I then have it read the draft to me, and I go through an iterative process of figuring out what ideas I like or don’t like.

I also use GPT to identify questions I need to ask within the chapters of my essay. My planned essay structure is: introduction, a chapter for the proposal, a chapter against the proposal, and a chapter of analysis and conclusion. By iteratively listening to GPT and reviewing my skeleton, I generated about 90 to 100 questions to analyse my sources with. From this I was able to produce what I think is a decent skeleton of an essay. All of this was happening while I was online every day, listening to the different lectures.

The lectures this week included Jeremy Maxwell. I really liked his style — much more conducive to learning. He comes across as relaxed and not as the “sage on the stage”. Rory purports not to be the sage on the stage, but in reality, he comes across like that. Jeremy doesn’t. His mannerisms worked very well, and he got us talking about interesting discussions on security. I got some good notes from his session for my essay on the Nature and Character of War.

We also had Jacob Westberg from the Swedish Defence Academy. His lecture was professionally delivered. I asked him a direct question about my DSS essay, where I argued that legitimacy is the centre of gravity for small states’ influence. He concurred with that, which was affirming.

Then we had a series of lectures from Dr Ian Speller on strategy. I found them very interesting. Not all of it was new to me, but his style I liked. He is clearly authoritative in his understanding and comprehension of the literature.

By contrast, I found Rory Finegan more difficult to listen to. He often seemed to be bluffing and unable to answer questions clearly. I would rate him about 4 out of 10. Jeremy I would give 8 out of 10 for style. Content is harder to judge at this stage, as he hasn’t gone deep yet. Ian Speller I would give 10 out of 10 for content, and I liked his style of delivery too.

My main academic focus this week was on trying to understand the difference between realism, liberalism and structural realism. I was reading Waltz’s book and started listening to John Mearsheimer. The main takeaway this week, however, was simply the workload — a huge amount of reading. I felt I did not have time to properly reflect on or absorb the readings.

I found the SRDs — the syndicate room discussions — very interesting. I realised that instead of just venting, I was trying to draw parallels and use the terminology we have been taught. I tried to link comments to authors I had read. However, I found it difficult because I was reading so much that I struggled to recall which ideas came from which specific paper.

I also chatted with Owen — OB — twice this week, on Wednesday and Thursday. On the first day he asked how I was getting on with the course, and I told him I was enjoying it. But the following night I texted him to say I had lied. In truth, I am not really enjoying the course. It is doing exactly what I did not want it to do: consuming a significant amount of my time.

The house is currently in a mess, with jobs half completed in every room. I am not happy about that. There are personal difficulties at home that the workload prevents me from addressing. I am not getting much time to exercise either. Much of this is my own fault, but I feel I am being sucked into the course against my better judgment — exactly what I did not want to happen.

⸻

Would you like me to also prepare Week 1 in the same style, so your portfolio will have a running set of raw transcripts you can draw from?
