\chapter{Literature}

\section{Introduction}Much of warfare conducted during the first two decades of the 21\textsuperscript{st}~century was heavily influenced by the IED. The threat and the response evolved continually. Crucially, techniques were rarely rendered obsolete. Instead, their effectiveness fluctuated over time. This evolutionary pattern accords with Krepinevich’s insistence that RMAs require organisational adaptation alongside technology \parencite{KREP_1992} and Metz’s observation of incremental rather than discontinuous change \parencite{METZ_2000,KREP_1994}. This is evidenced by the continued relevance of the Philosophy of IEDD, which endures as international doctrine since the 1970s  \parencite{COCHRANE_2012,DORD_2022}. An RMA would perhaps fundamentally transform the battlefield by rendering existing techniques and equipment permanently obsolete. This example may reflect what critics of the RMA thesis argue, that adaptation is cyclical and rarely results in true discontinuity. %[Consider integrating a brief reference to a key RMA theorist or critic who adopts a similar evolutionary view to ground the framing academically (e.g., Colin Gray or Steven Metz).]

 
%Krep's 1992 This article appears to foretell a lot from the future. the use of hybrid warfare by Russia and Iran to counter their clear inferiority in conventional forces. The requirement of agile and flat experimentation. The requirement of structures and people to implement the technological advancements. that older legacy systems can be used but in a different manner as newer systems take over the battle. the use of drones and long range precision fires instead of MBTs. that information is powera and that change in a bureacratic organisation can be very difficult. for example, the brits invented the tank and had thinkers who wanted to use it. but their system failed andinstead it wa sthe germans who could be agile enough to implement. that the use of civilian technolgoy is important and that the military can no longer practice a form of socialism where it owns all the tech. \parencite{KREP_1992}

\section{Foundational RMA Theories and Techno-Optimist Views}

Cohen suggested to consider the longevity of ``basic counting pieces of
military power"  \parencite{COHEN_1995}, echoing RCP. Discontinuous jumps in RCP aligns well with Krepinevich's 1992 description, whereby to disregard an RMA likely condemns your soldiers to slaughter. 

Krepinevich 1992's paper placed him as a seminal RMA thinker \nocite{KREP_1992}. He framed the RMA as more than just new platforms. It required agile, flat experimentation and organisation. He stressed that information itself would become power and military adoption of superior civilian technology. He also suggested that older systems could still play a role, but increasingly in support of new technologies such as long-range precision strike and drones which would displace armour as the centre of battle \nocite{KREP_1992}. %His later work reaffirmed that true revolutions require the co-evolution of technology, operational concepts, and institutions, rather than technology alone \parencite{KREP_1994}.%
 This sits alongside Cohen’s idea that RMAs function like hypotheses to be tested in war \parencite{COHEN_1996} and Owens’ vision of linked “systems of systems” reshaping force structure \parencite{OWENS_2002}. RMA optimists have felt that evolution in technology and thinking which were displayed during Desert Storm\index{Gulf War, 1990-1991} would clear the ``fog of war''  \parencite[p.~49]{ALACH_2008}. More recently, Brose argued that AI, autonomy and ubiquitous sensors will upend legacy concepts in the same way that smokeless powder once did, demanding a radically different force design built around swarms of cheap, expendable system \parencite{BROSE_2019}.

Their works suggest that Cohen and Krepinevich view the RMA concept as not just an abstract model within which to attempt to understand warfare. Instead an RMA is akin to a scientific hypothesis which can be tested. If one's inaction resulted in the `needless' slaughter of your soldiers, then one erroneously dismissed an RMA. It is clear from the literature, that proponents of the RMA framework do not simply see it as technological. Cultural, bureaucratic, financial, political and other factors influence to transform a technological invention to a military breakthrough. Metz (2000) similarly stresses that revolutions in military affairs are as much social, political and organisational as they are technological\nocite{METZ_2000}. The cultural context shapes whether new capabilities amount to true revolution.

\section{Critical and Skeptical Perspectives on RMA}

Krepinevich also recognised that technology on its own was insufficient. For him, invention without reorganisation was a dead end \parencite{KREP_1992}. This reinforces Betts’ warning that militaries often misuse technology if it is not embedded in doctrine and culture \parencite{BETTS_1996}. Krepinevich’s review of the Gulf War air campaign likewise underlined the limits of technology’s impact when political control and cultural factors remained intact \parencite{KREP_1996}. It also reflects Gray’s argument that strategic culture, not hardware, usually decides whether an apparent breakthrough becomes a genuine revolution \parencite{GRAY_2005}. Betts (1996) (and to a lesser extent Owens (2002)\nocite{OWENS_2002}) cautioned against concluding that the results of Desert Storm confirm the existence of an RMA\index{Gulf War, 1990-1991}. He assesses it as improbable that that the US is able to repeat the results of the 1990-1991 Gulf War in the future. Indeed he appears to have correctly predicted the US' inability to succeed against insurgencies it faced during GWoT. William Owens  similarly highlights that while technological advances offer extraordinary military effectiveness, organisational and cultural transformation lag behind and dilute its impact \parencite{OWENS_2002}. 

The skeptical tradition applies a kind of Occam’s razor to claims of revolution. Rather than positing epochal breaks, it stresses that continuity and incremental adaptation usually suffice to explain apparent change. As Alach argues, much of the supposed RMA discourse is rhetorical excess layered onto what are, in practice, evolutionary adjustments \parencite[.p~50-52]{ALACH_2008}. His conclusion is that there is an evolution in military affairs rather than a revolution. Rassler’s 2015 study of non-state drone innovation reinforces this skepticism, showing that much of the apparent novelty stems from incremental civilian adaptation rather than decisive military transformation”  \parencite{RASSLER_2015}.

\section{Doctrinal and Organisational Learning Perspectives}

The tank case is also an example of organisational learning and failure. Krepinevich showed that Britain had the concept but not the structures to exploit it \parencite{KREP_1992}. He later extended this logic, showing that historical patterns of military change confirm the decisive role of organisational adaptation over invention alone \parencite{KREP_1994}. This appears to align with Betts' description of military commanders approaching new technology with ``conservative progressivism  ''\parencite{BETTS_1996}.  Conversely inter-war Germany's circumstances allowed for a more flexible approach, resulting in transformation. Metz's (2000) and Owens (2002) articles accurately correlates recent defeat or a perception of weakness with openness to creativity\nocite{METZ_2000,OWENS_2002}. This was clearly the case for inter-war Germany.



Cohen's (1996) and Krepinevich's (1992) articles suggest that revising hierarchical structures is among the hardest tasks in realising an RMA. The recurring prescription is not mere acquisition/invention. It is often organisational change, faster decision cycles, decentralised command which realise the gain. Yet both tend to understate the cultural, political, budgetary and career incentives which impede revolutionary change\nocite{COHEN_1995,COHEN_1996,KREP_1992}. Owen's 2002 article clearly places him as an RMA optimist \nocite{OWENS_2002}. However, when reviewing Desert Storm\index{Operation Desert Storm|see{Gulf War, 1990-1991}}, he observed that apparent technological difficulties were, ``rooted in deeper differences of service culture, procedures, and operational concepts''. As Keller (2002) observed in his interviews with reform advocates, even within the Pentagon the frustration was less about technology itself than about a sclerotic culture that rewarded continuity over disruptive change\nocite{KELLER_2002}. The Stimson Center similarly noted that while UAS had become indispensable to U.S. commanders, their adoption was shaped less by radical reorganisation than by incremental integration into existing bureaucratic routines \parencite{STIMSON_2015}.

 Schaus and Johnson caution that UAS use also alters escalation dynamics, lowering thresholds without necessarily changing organisational cultures, which may create dangerous gaps between intent and perception” \parencite{SCHAUS_2018}. This is echoed by Krepinevich (1992).

%adam The lesson fits with Nagl’s claim that military culture decides whether adaptation happens (Nagl, 2002 NO -LINK) and with Fitzgerald’s description of doctrinal “amnesia” when lessons are ignored (Fitzgerald, 2013 - NO LINK). His work therefore links early RMA theory with later accounts of how culture and doctrine shape learning.

\section{AI, Autonomy, Mission Command and Contemporary Debates} 
In 2000, Metz  posited that historic commanders such as  or Guderian would likely have found the \gls{usalb} compatible with their operational style\index{AirLand Battle}\index{Doctrine!AirLand Battle}\index{Mission Command!AirLand Battle}\nocite{Metz_2000}. Central to this is the concept of mission command\index{Doctrine!Mission Command}. The tension between mission command's rapid operational tempo and the temptation of micro-management due to digital visibility is noteworthy.

Cohen warns of that technologies facilitating modern to commanders to perch ``cybernetically'' beside their troops in combat could have an insidious effect - undermining subordinate commanders \parencite{COHEN_1996}. He contrasts this with General Eisenhower and Field Marshall von Moltke lying on a sofa reading a book on the eve of battle. His skeptical characterisation of commanders prone to intervene indicates that Cohen is likely a proponent of mission command. Yet as Betts observes, new technology does not sharpen judgment by itself. Militaries often misinterpret or misuse innovations \parencite{BETTS_1996}. Together these cautions suggest that without cultural and organisational restraint, information systems may become instruments of centralised control rather than enablers of mission command. Alach (2008) concurs, stating that, ``mental evolution was as critical" as technological progress\nocite{ALACH_2008}.

As a proponent of mission command, the actions recounted by General Guderian are somewhat incongruent\index{Auftragstaktik| see{Mission Command}}\index{Mission Command!Auftragstaktik}\index{Guderian, Heinz}\index{Mission Command!Guderian, Heinz}. He described driving around the battlefield in his staff car to observe and intervene, at times confronting subordinates he believed were disobeying orders. One such episode involved SS Oberf\"uhrer  Dietrich\index{Dietrich, Sepp}\index{Mission Command!Guderian, Heinz}, whose apparent insubordination proved correct once Guderian examined the situation: ``I approved the decision taken by the commander on the spot'' \parencite[p.~117]{GUDERIAN_1952}. His preference for active oversight was further reflected in his praise for SS Gruppenf\"uhrer Paul Hausser\index{Hausser, Paul}\index{Mission Command!Hausser, Paul} \parencite[p.~73]{YEIDE_2011}. %General George S. Patton was similarly known for a hands-on style often criticised as ``micromanaging'' \parencite[p.~34]{ZALOGA_2010}\index{Patton, George S.}\index{Mission Command!Patton, George S.}.  

The tension between personal intervention and the philosophy of mission command is illustrated by Major Dick Winters\index{Winters, Dick}\index{Mission Command!Winters, Dick} during the Battle of Bastogne. Though instinct urged him to relieve Lieutenant Dike and lead Easy Company himself, Winters chose instead to uphold his broader responsibilities as battalion commander. His solution---appointing Lieutenant Ronald Speirs---proved decisive \parencite[p.~186]{WINTERS_2006}.  

Steven Metz argues that the USALB blended modern technology with Auftragstaktik and rapid tempo in ways Guderian or Patton would have recognised \parencite{METZ_2000}. More recently, Husain suggested that AI could compress the OODA loop to machine speed, accelerating conflict while still leaving space for the enduring principles of mission command \parencite{HUSAIN_2021}.


So it appears that Cohen underestimates the potential for technology to reinforce Auftragstaktik. His framing risks a reductive binary between passive commanders ``on the sofa” and intrusive commanders ``in the hatch”. Even if a commander’s judgment is superior, they cannot personally manage every decision across a dispersed and complex battlefield. What matters is leveraging subordinates’ initiative at scale, while reserving intervention for moments of crisis or opportunity. In this sense, the commander amplifies rather than replaces subordinate action. Hence, technology such as uncrewed systems and AI may enhance mission command. Krepinevich (1992) modified structures can enable technology to reinforce Auftragstaktik rather than undermine it\nocite{KREP_1992}. This highlights that the impact of technology is contingent less on the tools themselves than on organisational culture and doctrinal restraint. In favourable conditions, AI and uncrewed systems may extend, rather than erode, the practice of mission command. %Ochs' findings within multinational units are of note, wherein culture and interpretation can blunt mission command - despite the use of digital tools which would otherwise enhance it \parencite{OCHS_2020}.
 Authors such Alach, Gray, Betts (along with conditional optimists such as Cohen, Krepinevich and Metz) identify the incremental change/evolution combined and interactions between many systems to be a key consideration.

%Ochs finds that in multinational units, differences in culture and interpretation of Auftragstaktik often blunt the effectiveness of mission command, despite digital tools that could otherwise enhance decentralisation” (Oc
% get Gray and Kaldor's articles
\paragraph{Summary and Conclusion}\mbox{}\\
Krepinevich stands out as a bridge across the literature. He shared the techno-optimist belief that information, precision strike and new organisational forms could drive a revolution. Yet he also stressed that bureaucratic resistance and doctrinal inertia often prevent militaries from realising this potential. His recognition that invention without reorganisation fails, links early RMA theory to later institutional learning perspectives. In this sense, he links the optimism of Metz, Cohen and Owens with the caution of Betts, Alach and Gray. His insights anticipate contemporary debates on AI, autonomy and dual-use technology. This dual perspective is evident across his writings—from his broad historical analysis of past military revolutions \parencite{KREP_1994} to his tempered assessment of Desert Storm’s limited revolutionary character \parencite{KREP_1996}.